---
title: "Simple Bayesian Interim Looks"
author: "Vitaly Druker"
categories: [R, simulation, clinical trial design, bayesian]
bibliography: references.bib
---

## Notes

Conjugate normal prior:

```{r}
true_var <- 9
true_mean <- 1

sample_data <- rnorm(1e3, true_mean, sqrt(true_var))


prior_mu <- 0
prior_var <- 20



update_prior <- function(prior_mu, prior_var, new_data, true_var) {
    lik_mean <- mean(new_data)
    lik_var <- var(new_data)
    n <- length(new_data)

    # https://real-statistics.com/bayesian-statistics/bayesian-statistics-normal-data/conjugate-priors-normal-distribution/
    # new_var <- prior_var * true_var/(true_var + n * prior_var)
    # new_mu <- (prior_mu/prior_var + n*lik_mean/true_var)*new_var

    # https://people.eecs.berkeley.edu/~jordan/courses/260-spring10/lectures/lecture5.pdf
    new_var <- (1/prior_var + n/true_var) ^(-1)
    new_mu <- (prior_var * lik_mean + true_var * prior_mu)/(true_var/n + prior_var)

    list(mu = new_mu,
        var = new_var)

}

post_prob_gt <- function(prior_mu, prior_var, new_data, true_var, post_test = 0) {

    new_params <- update_prior(prior_mu, prior_var, new_data, true_var)

    pnorm(post_test, mean = new_params$mu, sqrt(new_params$var), lower.tail = FALSE)


}


post_prob_gt(
    prior_mu, 
    new_data = sample_data[1:100],
    true_var = true_var)
```

```{r}
power_analysis <- function(check_at = c(100, 200),
                           true_mean = 0,
                           prior_var = 20,
                           prior_mu = 0,
                           true_var = 49,
                           post_test = 0,
                           nsim = 1000) {
    names(check_at) <- paste0("L", seq_along(check_at))
    purrr::map(1:nsim, function(.s) {
        sample_data <- rnorm(max(check_at), true_mean, sqrt(true_var))
        purrr::map_dbl(check_at, function(.n) {
            post_prob_gt(
                prior_mu = prior_mu,
                prior_var = prior_var,
                new_data = sample_data[1:.n],
                true_var = true_var,
                post_test = post_test
            )
        })
    })
}

```


The results below show that type I error is inflated past te 
```{r}
null_example <- power_analysis(nsim = 1e4) |>
    dplyr::bind_rows()



mean(null_example$L2 > .975)
mean(null_example$L1 > .975)

mean(null_example$L1 > .975 | null_example$L2 > .975)

```

```{r}

null_power_sim <- power_analysis(nsim = 1e6, prior_var = 2) |>
    dplyr::bind_rows()


null_power_sim |>
    with(
        c(
            interim_success = mean(L1 > .975),
            final_success = mean(L2 > .975),
            either_success = mean(L1 > .975 | L2 > .975)
        )
    )


post_cutoffs <- seq(.975, 1, length.out = 10)

post_cutoffs_df <- expand.grid(L1 = post_cutoffs, 
            L2 = post_cutoffs)

post_cutoffs_df$either_success <- sapply(
    seq_len(nrow(post_cutoffs_df)), 
function(.r) {
    with(null_power_sim, {
        mean(L1 > post_cutoffs_df$L1[.r] | L2 > post_cutoffs_df$L2[.r])

    })
}

)


ggplot(post_cutoffs_df, aes(x = L1, y = L2, fill = either_success)) + 
geom_tile() +
geom_contour(aes(z = either_success, color = either_success < .025))




alter_power_sim <- power_analysis(
    nsim = 1e6,
    prior_var = 2,
    true_mean = 1
) 

alter_power_sim |>
    dplyr::bind_rows() |>
    with(
        c(
            interim_success = mean(L1 > .975),
            final_success = mean(L2 > .975),
            either_success = mean(L1 > .975 | L2 > .975)
        )
    )



power.t.test(
    n = 200,
    delta = 1,
    sd = sqrt(49),
    sig.level = 0.028227,
    type = "one.sample",
    alternative = "one.sided"
)
```

```{r}

# https://keaven.github.io/gsd-tech-manual/index.html
gsDesign::gsDesign(
    k = 2,
    test.type = 1,
    alpha = 0.028227,
    beta = 1 - 0.5412366,
    delta = 1/sqrt(49)

)

```

@ryan2020 shows that type I error is still an issue. Using skeptical prior may help. @kopp-schneider2020 shows that it's not possible to improve power with dynamic borrowing.

These two references were called out in @ryan2020 as recommending skeptical priors:

- Dmitrienko A, Wang M-D. Bayesian predictive approach to interim monitoring in clinical trials. Stat Med. 2006;25:2178--95. 28.
- Spiegelhalter DJ, Abrams KR, Myles JP. Bayesian approaches to clinical trials and health-care evaluation. Chichester: Wiley; 2004
