---
title: "Cutting with Trees"
output:
  html_document: default
  html_notebook: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(comment = NA)
knitr::opts_chunk$set(results = 'asis')
knitr::opts_chunk$set(tidy = TRUE)

library(dplyr)
library(tree)
library(ggplot2)
library(pander)
library(readr)

```

### Motivation  

As [I have mentioned before](http://www.westendstats.com/2016/11/conditional-mle-variance/), the most useful information statistics can provide us is a measure of uncertainty in a result. Arguably the most common way that researchers accomplish this is with [the p-value](http://blog.minitab.com/blog/adventures-in-statistics-2/how-to-correctly-interpret-p-values). Whatever your personal feelings about the use of p-values, it's important to understand and recognize specific instances when they can be misused. 
One way that I've seen this happen is while splitting a continuous 'control' variable into a categorical variable. While there can be [justification for using this method](https://www.ma.utexas.edu/users/mks/statmistakes/dividingcontinuousintocategories.html), it's generally [considered a bad idea](http://onlinelibrary.wiley.com/doi/10.1002/sim.2331/full).

Below, I will use a simple example to both show the issue with and the attraction to using regression trees to dichotomize a continuous control variable. Before we get to the simulation, let's imagine an example when a researcher might want to split a continuous variable. 

### Example  

Let's imagine a researcher is testing a therapy, recorded in the column `cat` as "A" or "B". The researcher is trying to understand if this particular therapy has an effect on a clinical outcome, such as hospital length of stay, recorded in column `val`. However, there is a third variable that may influence how well the treatment affects patients (such as age) recorded in `cont_cat`. The researcher wants build a multi-variate linear regression to control for `cont_cat`. However, the `cont_cat` data is complex, maybe it's easier to just split the data based off of the outcome variable. "This could take of non-linearity and call for a simpler interpretation," the researcher thinks. 

### Simulation  

While it sounds like an attractive proposition, it can get the researcher into trouble. The function below offers a simulation of what the researcher proposed:

```{r simulation_code}
sim_function <- function(samp_size = 1000, cut_ratio = .1, d = NULL) {
  # If data is not provided, simulate it
  if (is.null(d)){
    d <-
      data.frame(
        cat = rep(c('A', 'B'), samp_size/2),
        cont_cat = rnorm(samp_size),
        val = rlnorm(samp_size)
      )
  }else{
  # If data is provided, sample it
    d <- d[sample( 1:nrow(d), samp_size),]
  }
  
  # Calculate the optimal cut point
  r_cut <-
    tree(val ~ cont_cat,
         data = d,
         mindev = 0,
         mincut = samp_size*cut_ratio)$frame$splits[1, 'cutleft'] %>%
    substring(2) %>%
    as.numeric
  
  # Create high/low variable based off the cut above
  d$cut_var <-
    cut(
      d$cont_cat,
      breaks = c(min(d$cont_cat), r_cut , max(d$cont_cat)),
      include.lowest = T,
      labels = c('low', 'high')
    )
  
  # Fit linear models
  tree_mod <- lm(val ~ cat + cut_var, data = d)
  cont_mod <- lm(val ~ cat + cont_cat, data = d)
  
  # Return values signficant at alpha = .05
  c(
    "Dichotomized" = summary(tree_mod)$coefficients['cut_varhigh', 'Pr(>|t|)'] < .05,
    "Continuous" = summary(cont_mod)$coefficients['cont_cat', 'Pr(>|t|)'] < .05,
    "Cut Value" = r_cut
  )
}
``` 

I won't discuss the code above in detail, but the most important part is the call to the `tree` function (from the `tree` library). This function will find the optimal point to split the `cont_var` that best predicts `val`. The last rows output the results of testing for $\alpha = 0.05$ which means that our false positive rate should be controlled at 5%. The block of code below will check the significance form 1000 simulations to see if the false positive rates stays at that number.

```{r run_simulation, results = 'asis'}
full_sim <- sapply(1:1000, function(x)
  sim_function(1000, cut_ratio = .1))

out <- data.frame(`Precent Significant` = rowMeans(full_sim))

pander::pander(out, style = 'rmarkdown')
```

Oh oh... Look at that rate of finding a significant result with the `Dichotomized` method! It's 10x the prespecified $\alpha$ level! On the other hand, using the continuous value  keeps us at the prespecified false positive rate. 

This example clearly shows the issues that can come from an analysis that uses this technique. We know from the simulation that there is no relationship between the `cont_val` and the outcome variable `val`. However, in the extreme example above, we find a relationship _almost 50% of the time_.

### Real Data Simulation  

The code below illustrates just why people do this. You can see that it is easier to find a relationship (in the case below it does actually exist).

```{r ames_simulation}
data <- read_tsv("AmesHousing.txt")

d <- data %>% 
  filter(`House Style` %in% c('1Story','2Story'),
         `Gr Liv Area` < 4000) %>% 
  select(cat = `House Style`,
         cont_cat = `Lot Area`,
         val = SalePrice)

d$cat <- ifelse(d$cat == '1Story','A','B')

d$cont_cat <- scale(d$cont_cat)
d$val <- scale(d$val)


real_sim <- sapply(1:1000, function(x)
  sim_function(50, .1, d))

out <- data.frame(`Precent Significant` = rowMeans(real_sim))

pander::pander(out, style = 'rmarkdown')

```  

As you can see, we are able to detect an effect much more often when we dichotomize the variable. This shouldn't come as too much of a surprise as we are effectively sacrificing a low positive rate for a high specificity. As always, there is no such thing as a free lunch. 

As always, you can find all of the code for this post on [github](https://github.com/vitallish/westendstats/tree/master/2017/01)
