---
title: "Bayesian Power Trial"
author: "Vitaly Druker"
date: ""
categories: 
    - R
    - simulation
    - clinical trial design 
    - bayesian
    - power calculations
draft: false
bibliography: references.bib
---

## Overview of SPYRAL HTN-ON

Try to recreate power analysis in @böhm2020 - renal denervation trial that uses this method

-   Specifically the ON-MED group

-   Note that in the table below I transformed the standard errors found in table 3 of the publication for the pilot study.

    ```{r}
    parameter_assumptions <- tibble::tribble(
        ~trial,    ~arm,        ~basline_adjusted_mean, ~baseline_adjusted_sd, ~n,
        "pilot",   "treatment", -8.8,                   1.8*sqrt(36),          36,
        "pilot",   "control",   -1.8,                   1.8*sqrt(36),          36,
        "pivotal", "treatment", -6.8,                   12,                    NA,
        "pivotal", "control",   -1.8,                   12,                    NA
        
    )
    ```

Weibull discount functyion parameters that were used: Shape $k = 3$, scale $\lambda =0.25$

Interim analyses will happen at 175 and 220 subjects

Treatment effect defined by:

$\mu = \mu_t - \mu_c$

Trial success criteria:

$$
P(\mu <0 ) \gt .975
$$

Trial futility is made by imputation of remaining subjects and if

$$
P(\mu <0 ) \lt .05
$$

### Trial Performance Characteristics

- Overall trial pwer to detect treatment difference of -5 was 96%
- Type I error 3%
- Power at first and second interim looks was 89% and 94%


## Other publications

```{r}
library(bayesDP) # package that was used in the clinical trial
```

@haddad2017 for perspective from device community

Overview of different borrow strategies can be found here @viele2014. It didn't have details on how to implement dynamic borrowing in the way the @böhm2020 did.
