---
title: "Rare Categorical Variables"
output:
  html_notebook: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(comment = NA)

```

```{r libraries}
library(readr)
library(dplyr)
library(pander)
```

### Motiation  

I was working on the [Allstate Kaggle Competition](www.kaggle.com) and I ran into one particular problem over and over again as I was training/evaluating certain models. I was using cross-validation (5-fold) to approximate test error, and the evaluation kept failing because the model saw 'new' column values.

After digging into the data, I noticed that some columns in the data set contained _very_ rare factor values. Some might show up only once in the entire dataset! This becomes a problem for cross-validatiom because each fold does not contain the same column values.

Let's look at specific example. The table below shows the values in a column of a dataframe along with a label for the 'fold number' that we have placed that row in for cross-validation.

```{r create_example, echo = F}
set.seed(1)
samp_mat <- data.frame(Value = sample(c('A','B', 'C', 'D'),size = 20, prob = c(.3, .3, .3, .1), replace = T),
           Fold = rep_len(1:5, length.out = 20))

samp_mat %>% t() %>% pander

```
We can summarise this table to show the number each 'Value' appears in each fold:
```{r}

xtabs(~Value+Fold, data = samp_mat) %>% pander


```

We can clearly see that some column values are not represented in every fold. For example, A does not show up fold 3 and 4,


| Value | Fold Number |
|-------|-------------|
|A| 1|


### Simulation Comparing Observed and Expected Fisher Information  


#### Estimating $\hat{\theta}$ using MLE

```{r load_data}
train <- read_csv('data/train.zip', progress = F)

dim(train)
```

```{r cv_validity}
columnCounts <- function(x) {
  # Returns counts of unique values in a column
  if (!is.numeric(x) | is.factor(x)) {
    table(x)
  } else{
    ## ignore numeric columns
    NA
  }
}

filterLowCounts <- function(x, cutoff) {
  # Only returns values below a specific cutoff
  if (!is.na(x[1])) {
    ## x[1] to avoid warnings when vector
    out <- x[x < cutoff]
    if (length(out) == 0) {
      NA # returns NA if there is nothing available after cutoff
    } else{
      out
    }
  } else{
    x
  }
}



findRareValues <- function(x_df, cutoff, returnFiltered = T) {
  # filters out rows with rare factors
  column_values <- lapply(x_df, columnCounts)
  
  rare_values <-
    lapply(column_values, filterLowCounts, cutoff = cutoff)
  
  
  labeled_df <- mapply(function(cols_def, act_vals) {
      if (!is.na(cols_def)) {
        act_vals %in% names(cols_def)
      } else{
        NA
      }
    },
    rare_values,
    x_df)
    
  labeled_df <- labeled_df[!is.na(labeled_df)]
  if (length(labeled_df) == 0) {
    message("Nothing to remove")
    return(list(
      filtered_df = NA,
      rare_rows = NA,
      rare_values_df = NA
    ))
  }
  labeled_df <- simplify2array(labeled_df)
  
  rare_rows <- apply(labeled_df, MARGIN = 1, FUN = any)
  
  ## Create dataframe of columns/values
  rare_values <- rare_values[!is.na(rare_values)]
  rare_out <-
    data.frame(column = character(0),
               values = integer(0),
               ind = character(0))
  for (i in 1:length(rare_values)) {
    rare_out <- rbind(rare_out,
                      data.frame(column = names(rare_values[i]),
                                 stack(rare_values[[i]])))
    
    
  }
  
  message(paste('Removing', sum(rare_rows), 'rows.'))
  filtered_df <- if (returnFiltered) {
    x_df[!rare_rows,]
  } else {
    NA
  }
  
  list(
    filtered_df = filtered_df,
    rare_rows = which(rare_rows),
    rare_values_df = rare_out
  )
}


out <- findRareValues(train, 10, T)
length(out$rare_rows)
out <- findRareValues(out$filtered_df, 10, T)
length(out$rare_rows)

```

```{r create_cv_folds}

createCVFolds <- function(x_df, folds = 3, num_samples = 1, max.iter = Inf){
  
  samp_label <- rep_len(1:folds, length.out = nrow(x_df))
  output_df <- data.frame(row_num = 1:nrow(x_df))
  i <- 1
  union_intersect_diff <- c()
  
  while (ncol(output_df) < (num_samples + 1) & i < max.iter) {
    samp_set <- sample(samp_label)
    out <-
      by(x_df, samp_set, function(x) {
        names(na.omit(unlist(lapply(x, columnCounts))))
      }, simplify = T)
    
    intersect_list <- c()
    union_list <- c()
    
    for (j in out) {
      if (length(intersect_list) == 0) {
        intersect_list <- j
      }
      intersect_list <- intersect(intersect_list, j)
      union_list <- union(union_list, j)
    }
    cur_val <- length(union_list) - length(intersect_list)
    union_intersect_diff <- c(union_intersect_diff, cur_val)
    if (cur_val == 0) {
      output_df <- cbind(output_df, samp_set)
    }
    print(paste0('Try:', i, ' Diff:', cur_val))
    i <- i + 1
  }
  if(ncol(output_df)>1){
    names(output_df) <- c('row_num', 1:((ncol(output_df)-1)))
  }
  list(labeled_rows = output_df,
       difference = union_intersect_diff)
}

cv_folds <- createCVFolds(x_df = out$filtered_df,folds = 5, max.iter = 50)

 ```
```{r}
library(MASS)

fitdistr(cv_folds$difference,"chi-squared", start = list(df = 1, ncp = 1))
qchisq()

hist(cv_folds$difference)

```




### Further Reading  
1. [Computer Age Statistical Inference][efron_hastie] provided the initial reason for exploring this on page 48. Additionally, an older manuscript (1978!) from the same authors is available [for free online](https://www.stat.tamu.edu/~suhasini/teaching613/expected_observed_information78.pdf). See the graph on page 460 of the pdf.  
2. [Geyer, 2003][geyer_mle] provided some much needed details on how to calculate the fisher information in R. 
3. A step by step proof on how to calculate the Expected Fisher Information (warning: it's awful) can be [found here in the solution to (i)][expected_fisher]

[efron_hastie]: https://web.stanford.edu/~hastie/CASI/
[geyer_mle]: http://www.stat.umn.edu/geyer/5931/mle/mle.pdf
[expected_fisher]: http://wwwf.imperial.ac.uk/~das01/MyWeb/M3S3/Handouts/WorkedExample-Cauchy.pdf